Equalizer: A Mature Parallel Rendering Framework

### 2017-04-03 paperIEEE.tex
Filling in
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@3 db7c848b-7755-4df2-b763-8796797afa52

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
%\usepackage{tabularx}
\usepackage{amsmath}
%\usepackage{url}
\usepackage{subfigure}
\usepackage{multirow}
%\usepackage{teaser}

@@

%----------------------------------------------------------------------
%% Abstract section.
\abstract {
  We present the features, algorithms and system integration necessary to
  implement a parallel rendering framework usable in a wide range of real-world
  applications, based on the basic architecture and implementation of the Equalizer
  parallel rendering framework presented in \cite{EP:07}.
} % end of abstract

@@

\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Various Equalizer use cases: (a) immersive CAVE, (b) display wall and
    (c) scalable volume rendering.}
  \label{FIG_teaser}
\end{figure*}

@@

The early fundamental concepts of parallel rendering have been laid down in \cite{MCEF:94} and \cite{Crockett:97}. A number of domain specific parallel rendering algorithms and special-purpose hardware solutions have been proposed in the past, however, only few generic parallel rendering frameworks have been developed.

\subsubsection*{Domain specific solutions}

Cluster-based parallel rendering has been commercialized for off-line rendering (i.e. distributed ray-tracing) for computer generated animated movies or special effects, since the ray-tracing technique is inherently amenable to parallelization for off-line processing. Other special-purpose solutions exist for parallel rendering in specific application domains such as volume rendering \cite{LWMT:97,Wittenbrink:98,HSCSM:00,SL:02,GS:02,NSJLYZ:05} or geo-visualization \cite{VR:91,AG:95,LDC:96,JLMV:06}. However, such specific solutions are typically not applicable as a generic parallel rendering paradigm and do not translate to arbitrary scientific visualization and distributed graphics problems.

Recently in \cite{NC:07}, parallel rendering of hierarchical level-of-detail (LOD) data has been addressed and a solution specific to sort-first tile-based parallel rendering has been presented. While the presented approach is not a generic parallel rendering system, basic concepts presented in \cite{NC:07} such as load management and adaptive LOD data traversal can be carried over to other sort-first parallel rendering solutions.

\subsubsection*{Special-purpose architectures}

Traditionally,  high-performance real-time rendering systems have relied on an integrated proprietary system architecture, such as the SGI graphics super computers. These special-purpose solutions have become a niche product as their graphics performance does not keep up with off-the-shelf workstation graphics hardware and scalability of clusters. However, cluster systems need more sophisticated parallel graphics rendering libraries, such as the one proposed in this paper.

Due to its conceptual simplicity, a number of special-purpose image compositing hardware solutions for sort-last parallel rendering have been developed. The proposed hardware architectures include Sepia \cite {MHS:99a,sepia}, Sepia~2 \cite{LMSBHa:01,LMSBH:01}, Lightning~2 \cite{Stoll01}, Metabuffer \cite{Blanke00,Zhang01}, MPC Compositor \cite{Muraki01} and PixelFlow \cite{Molnar92, Eyles97}, of which only a few have reached the commercial product stage (i.e. Sepia~2 and MPC Compositor). However, the inherent inflexibility and setup overhead have limited their distribution and application support. Moreover, with the recent advances in the speed of CPU-GPU interfaces, such as PCI Express and other modern interconnects, combinations of software and GPU-based solutions offer more flexibility at comparable performance.

% FlowVR \cite{AGLLMRR:04}, not really related that much here

\subsubsection*{Generic approaches}

A number of algorithms and systems for parallel rendering have been developed in the past. On one hand, some general concepts applicable to cluster parallel rendering have been presented in \cite{Mueller:95,Mueller:97} (sort-first architecture), \cite{SZFLS:99,SFLS:00} (load balancing), \cite{SFL:01} (data replication), or \cite{CMF:05,CM:06} (scalability). On the other hand, specific algorithms have been developed for cluster based rendering and compositing such as \cite{AP:98}, \cite{CKS:02} and \cite{YYC:01,SMLAP:03}. However, these approaches do not constitute APIs and libraries that can readily be integrated into existing visualization applications, although the issue of the design of a parallel graphics interface has been addressed in \cite{Igehy98}.
%
Only few generic APIs and (cluster-) parallel rendering systems exist which include VR Juggler \cite{BJHMBC:01} (and its derivatives), Chromium \cite{HHNFAKK:02} (an evolution of \cite{Humphreys99,Humphreys00,HEBSEH:01}) and OpenGL Multipipe SDK \cite{JDBJBCER:04,BRE:05,MPK}.

VR Juggler \cite{BJHMBC:01,JBBC:98} is a graphics framework for virtual
reality applications which shields the application developer from the
underlying hardware architecture, devices and operating system. Its main
aim is to make virtual reality configurations easy to set up and use
without the need to know details about the devices and hardware
configuration, but not specifically to provide scalable parallel
rendering. Extensions of VR Juggler, such as for example ClusterJuggler
\cite{BC:03} and NetJuggler \cite{AGLMR:02}, are typically based on the
replication of application and data on each cluster node and basically
take care of synchronization issues, but fail to provide a flexible and
powerful configuration mechanism that efficiently supports scalable
rendering as also noted in \cite{SWNH:03}. The presented system is different from VR Juggler in that it
fully supports scalable parallel rendering such as sort-first and
sort-last task decomposition and image compositing, it provides more
flexible node configurations which for example allow specifying
arbitrary task decomposition and image compositing combinations as
simple compound layouts. Furthermore, it it is fully distributed which includes
support for network swap barriers (synchronization), distributed objects
as well as image compression and transmission. In contrast to VR
Juggler, Equalizer supports multiple rendering threads per process,
which is important for multi-GPU systems.

While Chromium \cite{HHNFAKK:02} provides a powerful and transparent
abstraction of the OpenGL API, that allows a flexible configuration of
display resources, its main limitation with respect to scalable
rendering is that it is focused on streaming OpenGL commands through a
network of nodes, often initiated from a single source. This has also been
observed in \cite{SWNH:03}. The problem
comes in when the OpenGL stream is large in size, due to not only
containing OpenGL calls but also the rendered data such as geometry
and image data. Only if the geometry and textures are mostly static
and can be kept in GPU memory on the graphics card, no significant
bottleneck can be expected as then the OpenGL stream is composed of a
relatively small number of rendering instructions. However, as it is
typical in real-world visualization applications, display and object
settings are interactively manipulated, data and parameters may change
dynamically, and large data sets do not fit statically in GPU memory but
are often dynamically loaded from out-of-core and/or multiresolution
data structures. This can lead to frequent updates not only of commands and
parameters wich have to be distributed but also of
the rendered data itself (geometry and texture), thus causing the OpenGL stream
to expand dramatically. Furthermore, this stream of function calls and
data must be packaged and broadcast in real-time over the network to
multiple nodes for each rendered frame. This makes CPU performance and
network bandwidth a more likely limiting factor. While preserving a minimally
invasive API, the novel proposed system is better aimed at scalability
as the actual data access is decentralized in the distributed rendering
clients.

The performance experiments in \cite{HHNFAKK:02} indicate that Chromium is working
quite well when the rendering problem is fill-rate limited.
This is due to the fact that the OpenGL commands and a non-critical
amount of rendering data can be distributed to multiple nodes without significant problems
and since the critical fill-rate work is then performed locally on the graphics hardware.

Chromium also provides some facilities for parallel application
development, namely a sort-last, binary-swap compositing SPU and an
OpenGL extension providing synchronization primitives, such as a barrier
and semaphore. It leaves other problems, such as configuration, task
decomposition as well as process and thread management unaddressed, thus
making the development of parallel OpenGL applications harder than with
Equalizer. Parallel Chromium applications tend to be written for one
specific parallel rendering use case, such as for example the sort-first distributed
memory volume renderer \cite{BHPB:03} or the sort-last parallel
volume renderer raptor \cite{Raptor}. We are not aware of a generic Chromium-based
application using many-to-one sort-first or stereo decompositions. This is
another difference to Equalizer which provides a much more flexible
task decomposition configuration. Applications written once for Equalizer can
easily be run in any different task decomposition mode and for any physical
display configuration without any changes to the application itself.
While Equalizer provides an abstraction of all entities of the rendering pipeline
(see Sections~\ref{SEC_architecture} and~\ref{SEC_development}), Chromium's
infrastructure is primarily the compositing stage.

OpenGL Multipipe SDK (MPK) \cite{BRE:05} implements an effective
parallel rendering API for a shared memory multi-CPU/GPU system. It is
similar to IRIS Performer \cite{RH:94} in that it handles multi-pipe
rendering by a lean abstraction layer via a conceptual callback
mechanism, and that it runs different application tasks in
parallel. However, MPK is not designed nor meant for rendering nodes
separated by a network. MPK focuses on providing a parallel rendering
framework for a single application, parts of which are run in parallel
on multiple rendering channels, such as the culling, rendering and final
image compositing processes. Compared to MPK, Equalizer supports a
fully distributed parallel rendering paradigm and features a more flexible
task decomposition approach.

@@

@@

Canvas, Layout, Observer

multi-view with runtime switching,

2D overlays

Swap Barriers

\subsection{Auto-Configuration}

hwsd with VirtualGL support

@@

Equalizer are one addition to compound trees. They modify parameters of their
respective subtree at runtime to optimize one aspect of the decomposition.

@@

\subsection{Livre}
\subsection{RTT Deltagen}
\subsection{RTNeuron}

@@

>>>>

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
%\usepackage{tabularx}
\usepackage{amsmath}
%\usepackage{url}
\usepackage{subfigure}
\usepackage{multirow}

@@

%----------------------------------------------------------------------
%% Abstract section.
\abstract{ We present the features, algorithms and system integration necessary
  to implement a parallel rendering framework usable in a wide range of
  real-world research and industry applications, based on the basic architecture
  and implementation of the Equalizer parallel rendering framework presented in
  \cite{EMP:09}.  } % end of abstract

@@

\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
    untracked views running RTT DeltaGen for virtual car usability studies (c)
    Cave2 running a molecular visualization build using Omegalib.}
  \label{FIG_teaser}
\end{figure*}

@@

\cite{EMP:09} summarized the work in parallel rendering at the time of
publication. In the following we will present the related work published since
this publication.

\cite{EEP:11} cross-segment load balancing

\cite{DK:11}: Framework to develop apps for multitile, thus targeted
specifically for tiled-wall visualization systems: grid configuration,
scalability (?), multiple work sessions, multiuser events. Master-slave approach
like Eq. with grid nodes running instances of the application. Is maybe more
like a GLUT and window replacement, or enhancement to deal with the OpenGL
contexts and events

\cite{CKP:12} tiled display wall virtual environment

\cite{TBD} DisplayCluster

\cite{NHM:11} ClusterGL?

Omegalib

@@

In this section we present features motivated by real-world application use
cases, i.e., new functionalities rather then performance improvements. We
motivate the use case, explain the architecture and integration into our
parallel rendering framework, and, where applicable, show the steps needed to
use this functionality in applications.

@@

Real-world visualization setups are often complex, and having an abstract
representation of the display system can simplify the configuration
process. Real-world applications often have the need to be aware of spatial
relationship of the display setup, for example to render 2D overlays or to
configure multiple views on a tiled display wall.

We addressed this need through a new configuration section interspersed between
the node/pipe/window/channel hardware resources and the compound trees
configurating the resource usage for parallel rendering.

A typical installation consists of one projection canvas, which is one
aggregated projection surface, e.g., a tiled display wall or a CAVE. Desktop
windows are considered a canvas. Each canvas is made of one or more segments,
which are the individual outputs connected to a display or projector. Segments
can be planar or non-planar to each other, and can overlap or have gaps between
each other. A segment is referencing a channel, which defines the output area of
this segment, e.g., on a DVI connector connected to a projector.

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. A canvas may have a software or hardware swap barrier,
which will synchronize the rendering of all contributing GPUs.

On each canvas, the application can display one or more views. A view is a view
on a model, in the sense used by the MVC pattern. The view class is used by
Equalizer applications to define view-specific data for rendering, e.g., a
scene, viewing mode or camera. The application process manages this data, and
the render clients receive it for rendering.

A layout groups one or more views which logically belong together. A layout is
applied to a canvas. The layout assignment can be changed at run-time by the
application. The intersection between views and segments defines which output
channels are available, and which frustum they should use for rendering. These
output channels are then used as destination channels in a compound. They are
automatically created during configuration.

A view may have a frustum description. The view's frustum overrides frusta
specified at the canvas or segment level. This is used for non-physically
correct rendering, e.g., to compare two models side-by-side on a tiled display
wall. If the view does not specify a frustum, the corresponding destination
channels will use the physically correct sub-frustum resulting from the
view/segment intersection.

An observer looks at one or more views. It is described by the observer position
in the world and its eye separation. Each observer will have its own stereo
mode, focus distance and frame loop (framerate). This allows to have untracked
views and multiple tracked views, e.g., two HMDs, in the same application.

\subsection{Automatic Configuration}

Automatic configuration implements the discovery of local and remote resources
as well as the creation of typical configurations using the discovered resources
at application launch time.

The discovery is implemented in a separate library, hwsd (HardWare Service
Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
WGL windowing systems, as well as network interfaces on Linux, Mac OS X and
Windows. Furthermore, it detects the presence of VirtualGL to allow optimal
configuration of remote visualization clusters. The resources can be discovered
on the local workstation, and through the help of a simple daemon using the
zeroconf protocol, on a set of remote nodes within a visualization cluster. A
session identifier may be used to support multiple users on a single cluster.

The Equalizer server uses the hwsd library to discover local and remote
resources when an hwsd session name instead of a \textsf{.eqc} configuration
file is provided. A set of standard decomposition modes is configured, which can
be selected through activating the corresponding layout.

This versatile mechanism allows non-experts to configure asnd profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to Equalizer application developers.

@@

Equalizer are an addition to compound trees. They modify parameters of their
respective subtree at runtime to optimize one aspect of the decomposition.

@@

\subsection{Livre}
\subsection{RTT Deltagen}
\subsection{RTNeuron}
\subsection{Omegalib}

@@

\subsection{Object Distribution}

Distribute large ply model to 1..16 nodes using TCP, IB and RSP

\subsection{Decomposition Modes}

Two graphs: Time to render a 4k, 256-step MSAA image of largest ply model and
Livre using 1..16 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
tiles, chunks, pixel, subpixel)
